---
title: "Introduction to dfms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to dfms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  cache = TRUE,
  fig.width = 7,
  fig.height = 5,
  comment = "#>"
)
```

*dfms* provides a user friendly and computationally efficient approach to estimate linear Gaussian Dynamic Factor Models in R. The package is not geared at any specific application, and can be used for dimensionality reduction, forecasting and nowcasting systems of time series. The use of the package is facilitated by a comprehensive set of methods to explore/plot models and extract results.

```{r setup, message=FALSE}
library(dfms)
library(xts)
library(magrittr)
```

This vignette walks through the main features of the package using first a simple, single frequency dataset of monthly time series, followed by some suggestions on how to deal with mixed frequencies. The data provided in the package in *xts* format is taken from Banbura and Modugno (2014)^[Banbura, M., & Modugno, M. (2014). Maximum likelihood estimation of factor models on datasets with arbitrary pattern of missing data. *Journal of Applied Econometrics, 29*(1), 133-160.], henceforth BM14, and covers the Euro Area from January 1980 through September 2009.

```{r}
dim(BM14_M)
range(index(BM14_M))
head(colnames(BM14_M))
plot(scale(BM14_M), lwd = 1)
```

The data frame `BM14_Models` provides information about the series^[Both about the monthly ones and quarterly series contained in `BM14_Q`. The order of rows in `BM14_Models` matches the column order of series in `merge(BM14_M, BM14_Q)`.], and the various models estimated by BM14. 

```{r}
head(BM14_Models, 3)

# Using only monthly data
BM14_Models_M <- subset(BM14_Models, freq == "M")
```

Prior to estimation, all data is differenced by BM14, and some series are log, differenced, as indicated by the `log_trans` column in `BM14_Models`. In general, *dfms* uses a stationary Kalman Filter with time-invariant system matrices, and therefore expects data to be stationary. Data is also scaled and centered^[Data must be scaled and centered because the Kalman Filter has no intercept term.] in the main `DFM()` function, thus this does not need to be done by the user. 

```{r}
# log-transforming and first-differencing the data
BM14_M[, BM14_Models_M$log_trans] %<>% log()
BM14_M_diff = diff(BM14_M)
plot(scale(BM14_M_diff), lwd = 1)
```

Before estimating a model, the `ICr()` function can be applied to determine the number of factors. It computes 3 information criteria proposed in Bai and NG (2002)^[Bai, J., Ng, S. (2002). Determining the Number of Factors in Approximate Factor Models. *Econometrica, 70*(1), 191-221. <doi:10.1111/1468-0262.00273>], whereby the second criteria generally suggests the most parsimonious model. 

```{r}
ic = ICr(BM14_M_diff)
print(ic)
plot(ic)
```

Another option is to use a Screeplot to gauge the number of factors by looking for a kink in the plot. A mathematical procedure for finding the kink was suggested by Onatski (2010)^[Onatski, A. (2010). Determining the number of factors from empirical distribution of eigenvalues. *The Review of Economics and Statistics, 92*(4), 1004-1016.], but this is currently not implemented in `ICr()`. 

```{r}
screeplot(ic)
```

Based on both the information criteria and the Screeplot, I gauge that a model with 4 factors should be estimated, as factors, 5, 6 and 7 do not add much to the explanatory power of the model. Next to the number of factors, the lag order of the factor-VAR of the transition equation should be estimated (the default is 1 lag). This can be done using the `VARselect()` function from the *vars* package, with PCA factor estimates reported by `ICr()`.

```{r}
# Using vars::VARselect() with 4 principal components to estimate the VAR lag order
vars::VARselect(ic$F_pca[, 1:4])
```

The selection thus suggests we should estimate a factor model with ` r = 4` factors and `p = 3` lags^[Some authors like Doz, Giannone, and Reichlin (2012) additionally allow the number of transition error processes, termed 'dynamic factors' and given an extra parameter `q`, to be less than the number of 'static factors' `r`. I find this terminology confusing and the feature not very useful for most practical purposes, thus I have not implemented it in *dfms*.]. Before estimating the model I note that *dfms* does not deal with seasonality in series, thus it is recommended to also seasonally adjust data, e.g. using the *seasonal* package before estimation. BM14 only use seasonally adjusted series, thus this is not necessary with the example data provided. Estimation can then simply be done using the `DFM()` function with parameters `r` and `p`^[Users can also choose from two different implementations of the EM algorithm using the argument `em.method`. The default is `em.method = "auto"`, which chooses the modified EM algorithm proposed by BM14 for missing data if `anyNA(X)`, and the plain EM implementation of Dos, Giannone and Reichlin (2012), henceforth DGR12, otherwise. The basline implementation of DGR12 is typically more than 2x faster than BM14's method, and can also be used with data that has a few random missing values (< 5%), but gives wrong results with many and systematically missing data, such as ragged edges at the beginning of end of the sample or series at different frequencies. With complete datasets, both `em.method = "DGR"` and `em.method = "BM"` gives identical factor estimates, and in this case `em.method = "DGR"` should be used.]. 

```{r}
# Estimating the model with 4 factors and 3 lags using BM14's EM algorithm
model1 = DFM(BM14_M_diff, r = 4, p = 3)
print(model1)
plot(model1)
```

The model can be investigated using `summary()`, which returns an object of class 'dfm_summary' containing the system matrices and summary statistics of the factors and the residuals in the measurement equation, as well as the R-Squared of the factor model for individual series. The print method automatically adjusts the amount of information printed to the data size. For large databases with more than 40 series, no series-level statistics are printed. 

```{r}
dfm_summary <- summary(model1)
print(dfm_summary) # Large model with > 40 series: defaults to compact = 2

# Can request more detailed printouts
# print(dfm_summary, compact = 1)
# print(dfm_summary, compact = 0) 
```

Another way to examine the factor model visually is to plot the EM factor estimates against PCA and Two-Step estimates following Doz, Giannone and Reichlin (2011)^[Doz, C., Giannone, D., & Reichlin, L. (2011). A two-step estimator for large approximate dynamic factor models based on Kalman filtering. *Journal of Econometrics, 164*(1), 188-205.], there the Kalman Filter and Smoother is run only once. Both types of estimates are also returned b `DFM()` in EM estimation and can be visualized with `plot.dfm`.

```{r}
plot(model1, method = "all", type = "individual")
```







